AusTraits -- a curated plant trait database for the Australian flora

### 2020-11-10 ms.Rmd
Update author lists
fix minor grammar, quotes in tables

We introduce the AusTraits database - a compilation of measurements of plant traits for taxa in the Australian Flora (hereafter AusTraits). AusTraits synthesises data on `r austraits$traits$trait_name %>% n_distinct()` traits across `r austraits$traits$taxon_name %>% n_distinct()` taxa from field campaigns, published literature, taxonomic monographs, and individual taxa descriptions. Traits
vary in scope from physiological measures of performance (e.g. photosynthetic gas exchange, water-use efficiency) to morphological
parameters (e.g. leaf area, seed mass, plant height) which link to aspects of ecological variation. AusTraits contains curated and
harmonised individual-, species- and genus-level observations coupled to, where available, contextual information on site
properties. This data descriptor provides information on version  `r austraits$build_info$version` of AusTraits which contains data for `r austraits$traits %>% nrow()` trait-by-taxa combinations. We envision AusTraits as an on-
going collaborative initiative for easily archiving and sharing trait data to increase our collective understanding of the Australian flora.

@@

Species traits are essential metrics for comparing ecological strategies in plants arrayed across environmental space or evolutionary lineages \citep{diaz2016global, kunstler2016plant, zanne2014three, cornwell2014functional}. Broadly, a trait is any measurable property of a plant capturing aspects of its structure or function \citep{chapin1993evolution, adler2014functional, diaz1998plant,violle2007let}. Traits thereby provide useful indicators of species behaviours in communities and ecosystems, regardless of their taxonomy \citep{violle2007let, funk2017revisiting}. Through global initiatives the volume of available trait information for plants has grown rapidly in the last two decades \citep{Kattge2011try, Kattge-2020}. However, the geographic coverage of trait observations across the globe is patchy, limiting detailed analyses of trait variation and diversity in some regions.

One such region is Australia; a continent with a flora of c. 26,000 native higher-plant species \citep{council2010australian}. While significant investment has been made in curating and digitising herbarium collections and observation records in Australia over the last two decades (e.g. The Australian Virtual Herbarium houses ~7 million specimen occurrence records; https://avh.ala.org.au), no complementary resource yet exists for consolidating information on plant traits. Moreover, relatively few Australian species are represented in the leading global databases. For example, the international TRY database \citep{Kattge-2020} has observations for only 3830 Australian species across all collated traits. This level of species coverage limits our ability to use traits to understand and ultimately manage Australian vegetation \citep{Kissling_2018}. While initiatives such as TRY \citep{Kattge-2020} and the Open Traits Network \citep{Gallagher-2020} are working towards global synthesis of trait data a stronger representation of Australian plant taxa in these efforts is essential given the high richness and endemicity of this continental flora.

@@

As an evolving data product, successive versions of AusTraits are being released, containing updates and corrections. Versions are labeled using semantic versioning to indicate the change between versions \citep{Falster-2019}. Static versions of the AusTraits, including version`r austraits$build_info$version` used in this descriptor, are available on the project website (https://github.com/traitecoevo/austraits) and Zenodo (Data Citation 1). The latest data can also be downloaded directly from the project website. As validation (see Technical Validation, below) and data entry is ongoing, users are recommended to pull data from the static releases, to ensure results in their downstream analyses remain consistent as the database is updated.

@@

There were substantial differences in coverage among different tissues and trait types, also with respect to number of georeferenced points (Figure 2).The most common traits are nongeoreferenced records from floras. Yet, geo-referenced records were available in several traits for more than 10% of the flora (Figure 2a).

@@

Third, are providing a pathway for user feedback. As AusTraits is a community resource and we encourage engagement from users on maintaining the quality and usability of the dataset. As such, we welcome reporting of possible errors, as well as additions and edits to the online documentation for AusTraits that make using the existing data, or adding new data, easier for the community.

@@

This document provides information on the structure of AusTraits and corresponds to version `r austraits$definitions$austraits$elements$build_info$elements$version$value` of the dataset.

@@

![Number of taxa with trait records by plant tissue and trait category, for data that is  (A.) geo-referenced, and (B.) not geo-referenced. ](figures/austraits_fig_composition.pdf)

@@

![Coverage of georeferenced trait records across Australian climatic and geographic space for traits in different categories. (a.) AusTraits’ sites (orange) within Australia’s precipitation-temperature space (dark-grey) superimposed upon Whittaker’s major biomes diagram. Climate data were extracted from Worldclim \citep{Fick-2017}.(B.) Locations of geo-referenced records for different plant tissues. ](figures/austraits_fig_biome_map.png)

@@

```{r}
tibble(
  Element = names(austraits$definitions$austraits$elements), 
  Contents = sapply(austraits$definitions$austraits$elements, "[[", "description") %>% gsub("`", "’", ., fixed=TRUE)
  ) %>% 
  knitr::kable("latex",  booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("striped")) %>%
  kableExtra::column_spec(1, width = "15em") %>%
  kableExtra::column_spec(2, width = "30em") %>% 
  str_replace_all(fixed("gray!6"), fixed("gray!20")) %>%
   writeLines()
```

@@

    elements$elements %>%
    list1_to_df() %>%
    knitr::kable("latex",  booktabs = TRUE) %>% 
    kableExtra::kable_styling(latex_options = c("striped")) %>%
    kableExtra::column_spec(1, width = "15em") %>%
    kableExtra::column_spec(2, width = "30em") %>%
    str_replace_all(fixed("gray!6"), fixed("gray!20")) %>%
    writeLines()

@@

```{r, results="asis"}
austraits$definitions$value_type$values %>% 
  list1_to_df()  %>%
  knitr::kable("latex",  booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("striped")) %>%
  kableExtra::column_spec(1, width = "15em") %>%
  kableExtra::column_spec(2, width = "30em") %>%
  str_replace_all(fixed("gray!6"), fixed("gray!20")) %>%
  writeLines()
```

>>>>

We introduce the AusTraits database - a compilation of measurements of plant traits for taxa in the Australian flora (hereafter AusTraits). AusTraits synthesises data on `r austraits$traits$trait_name %>% n_distinct()` traits across `r austraits$traits$taxon_name %>% n_distinct()` taxa from field campaigns, published literature, taxonomic monographs, and individual taxa descriptions. Traits
vary in scope from physiological measures of performance (e.g. photosynthetic gas exchange, water-use efficiency) to morphological
parameters (e.g. leaf area, seed mass, plant height) which link to aspects of ecological variation. AusTraits contains curated and
harmonised individual-, species- and genus-level observations coupled to, where available, contextual information on site
properties. This data descriptor provides information on version  `r austraits$build_info$version` of AusTraits which contains data for `r austraits$traits %>% nrow()` trait-by-taxa combinations. We envision AusTraits as an on-
going collaborative initiative for easily archiving and sharing trait data to increase our collective understanding of the Australian flora.

@@

Species traits are essential metrics for comparing ecological strategies in plants arrayed across environmental space or evolutionary lineages \citep{diaz2016global, kunstler2016plant, zanne2014three, cornwell2014functional}. Broadly, a trait is any measurable property of a plant capturing aspects of its structure or function \citep{chapin1993evolution, adler2014functional, diaz1998plant,violle2007let}. Traits thereby provide useful indicators of species' behaviours in communities and ecosystems, regardless of their taxonomy \citep{violle2007let, funk2017revisiting}. Through global initiatives the volume of available trait information for plants has grown rapidly in the last two decades \citep{Kattge2011try, Kattge-2020}. However, the geographic coverage of trait observations across the globe is patchy, limiting detailed analyses of trait variation and diversity in some regions.

One such region is Australia; a continent with a flora of c. 26,000 native higher-plant species \citep{apc-2020}. While significant investment has been made in curating and digitising herbarium collections and observation records in Australia over the last two decades (e.g. The Australian Virtual Herbarium houses ~7 million specimen occurrence records; https://avh.ala.org.au), no complementary resource yet exists for consolidating information on plant traits. Moreover, relatively few Australian species are represented in the leading global databases. For example, the international TRY database \citep{Kattge-2020} has observations for only 3830 Australian species across all collated traits. This level of species coverage limits our ability to use traits to understand and ultimately manage Australian vegetation \citep{Kissling_2018}. While initiatives such as TRY \citep{Kattge-2020} and the Open Traits Network \citep{Gallagher-2020} are working towards global synthesis of trait data a stronger representation of Australian plant taxa in these efforts is essential given the high richness and endemicity of this continental flora.

@@

As an evolving data product, successive versions of AusTraits are being released, containing updates and corrections. Versions are labeled using semantic versioning to indicate the change between versions \citep{Falster-2019}. Static versions of the AusTraits, including version `r austraits$build_info$version` used in this descriptor, are available on the project website (https://github.com/traitecoevo/austraits) and Zenodo \citep{falster_2019}. The latest data can also be downloaded directly from the project website. As validation (see Technical Validation, below) and data entry is ongoing, users are recommended to pull data from the static releases, to ensure results in their downstream analyses remain consistent as the database is updated.

@@

There were substantial differences in coverage among different tissues and trait types, also with respect to number of geo-referenced points (Figure 2).The most common traits are non geo-referenced records from floras. Yet, geo-referenced records were available in several traits for more than 10% of the flora (Figure 2a).

@@

Third, ww provide a pathway for user feedback. AusTraits is a community resource and we encourage engagement from users on maintaining the quality and usability of the dataset. As such, we welcome reporting of possible errors, as well as additions and edits to the online documentation for AusTraits that make using the existing data, or adding new data, easier for the community. Feedback can be posted as an issue directly at the project website: https://github.com/traitecoevo/austraits.build/issues/new/.

@@

This document provides information on the structure of AusTraits and corresponds to version `r austraits$build_info$version` of the dataset.

@@

![Number of taxa with trait records by plant tissue and trait category, for data that is  (A.) geo-referenced, and (B.) not geo-referenced. Many records without a geo-reference come from botanical collections, such as floras. ](figures/austraits_fig_composition.pdf)

@@

![Coverage of geo-referenced trait records across Australian climatic and geographic space for traits in different categories. (a.) AusTraits’ sites (orange) within Australia’s precipitation-temperature space (dark-grey) superimposed upon Whittaker’s major biomes diagram. Climate data were extracted from Worldclim \citep{Fick-2017}.(B.) Locations of geo-referenced records for different plant tissues. ](figures/austraits_fig_biome_map.png)

@@

```{r}
clean_quotes <- function(txt) {
  str_replace_all(txt, "`([a-zA-Z_ 0-9()\\.\\\\_]+)`", " texttt{\\1}") %>% 
  str_replace_all(fixed(" texttt"), fixed("\\texttt")) 
}

tibble(
  Element = names(austraits$definitions$austraits$elements), 
  Contents = sapply(austraits$definitions$austraits$elements, "[[", "description")
  ) %>% 
  knitr::kable("latex",  booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("striped")) %>%
  kableExtra::column_spec(1, width = "15em") %>%
  kableExtra::column_spec(2, width = "30em") %>% 
  str_replace_all(fixed("gray!6"), fixed("gray!20"))  %>%
    clean_quotes() %>%
  writeLines()
```

@@

    elements$elements %>%
    list1_to_df() %>%
#    mutate(value = clean_quotes(value)) %>%
    knitr::kable("latex",  booktabs = TRUE) %>% 
    kableExtra::kable_styling(latex_options = c("striped")) %>%
    kableExtra::column_spec(1, width = "15em") %>%
    kableExtra::column_spec(2, width = "30em") %>%
    str_replace_all(fixed("gray!6"), fixed("gray!20"))  %>%
    clean_quotes() %>%
    writeLines()

@@

```{r, results="asis"}
austraits$definitions$value_type$values %>% 
  list1_to_df()  %>%
  knitr::kable("latex",  booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("striped")) %>%
  kableExtra::column_spec(1, width = "15em") %>%
  kableExtra::column_spec(2, width = "30em") %>%
  str_replace_all(fixed("gray!6"), fixed("gray!20")) %>%
  clean_quotes() %>%
  writeLines()
```
